---
title: "Spherical Geometry"
author: "Edzer Pebesma; OpenGeoHub summer school, Sept 1-3 2021"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
---

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Create a set of 1000000 (1e6) points, preferably spread (approximately) regularly over the Earth’s surface

O.K., we do a bit less.

```{r}
library(sf)
library(stars)
library(sp)
st_bbox() |>
  st_as_sfc() -> bb
as(bb, "Spatial") |>
  spsample(n = 1000, type = "Fibonacci") |> # https://arxiv.org/pdf/0912.4540.pdf
  st_as_sfc() -> x
plot(x)
plot(bb, add = TRUE)
```

Or, in orthographic:

```{r}
par(mfrow=c(2,2), mar = rep(0, 4))
plot(st_transform(x, st_crs("+proj=ortho +lon_0=45 +lat_0=-45")), pch = 3)
plot(st_transform(x, st_crs("+proj=ortho +lon_0=0 +lat_0=-45")), pch = 3)
plot(st_transform(x, st_crs("+proj=ortho +lon_0=-30 +lat_0=35")), pch = 3)
plot(st_transform(x, st_crs("+proj=ortho +lon_0=-90 +lat_0=-15")), pch = 3)
```

Some more points:

```{r}
as(bb, "Spatial") |>
  spsample(n = 1e5, type = "Fibonacci") |>
  st_as_sfc() -> x
```

## Subset this dataset for points that are on land

```{r}
library(rnaturalearth)
co = ne_countries(returnclass = "sf") |> st_make_valid()
x = x[co]
plot(x)
```

## Generate for the entire Earth, ...

... on a regular grid, densities of these land points (also
called intensity: number of points per unit area), using different
bandwidths (e.g. 50, 100, 200, 500 and 1000 km)

Create a function that takes a point set and a bandwidth as input
arguments and that returns a global grid with point densities

```{r}
(grd = st_as_stars(st_bbox(), nx = 90, ny = 45)) # creates 4-degree global grid
```

Slow:
```{r eval=FALSE}
calc_dens = function(pts, grd, dist) {
		buf = st_buffer(pts, dist)
		a = st_area(buf) |> units::set_units(km^2) |> units::drop_units()
		lengths(st_intersects(grd, buf)) / a
}
grd$d1000 = calc_dens(x, st_as_sfc(grd, as_points = TRUE), units::set_units(1000, km))
plot(grd["d1000"])
title("points per square km")
```

Faster:
```{r}
calc_dens = function(pts, grd, dist) {
		pt = st_sfc(st_point(c(0,0)), crs = 4326)
		a = st_area(st_buffer(pt, dist)) |> units::set_units(km^2) |> units::drop_units()
		lengths(st_is_within_distance(grd, pts, dist)) / a
}
grd$d100 = calc_dens(x, st_as_sfc(grd, as_points = TRUE), units::set_units(100, km))
grd$d200 = calc_dens(x, st_as_sfc(grd, as_points = TRUE), units::set_units(200, km))
grd$d500 = calc_dens(x, st_as_sfc(grd, as_points = TRUE), units::set_units(500, km))
grd$d1000 = calc_dens(x, st_as_sfc(grd, as_points = TRUE), units::set_units(1000, km))
grd$values = NULL
plot(merge(grd))
```

# Visualise these maps with densities using an orthographic projection, from several different perspectives

```{r}
par(mfrow=c(2,2), mar = c(0, 0, 1.1, 0))
sf = st_as_sf(grd)
ne = function(x, lat, lon) { 
		crs = st_crs(paste0("+proj=ortho +lon_0=", lon, " +lat_0=", lat))
		pt = st_sfc(st_point(c(lon, lat)), crs = 4326)
		bu = st_buffer(pt, units::set_units(9800000, m))
		x = st_transform(st_intersection(x, bu), crs);
		plot(x, key.pos = NULL, border = NA, reset = FALSE)
		plot(st_transform(st_intersection(st_geometry(co), bu), crs), add = TRUE)
}
ne(sf[4], 45, -45)
ne(sf[4], -30, 45)
ne(sf[4], -45, -135)
ne(sf[4], 90, 0)
```

## Create a spinning globe (animated) plot of the densities

## (for bonus points) build in the option to choose different kernel functions for density computation

## (for bonus points) let the densities not spill over in water areas, i.e. compute them only over land and take edge effects into account

This only clips the maps above, it doesn't compute taking water into account:
```{r}
plot(merge(grd)[co])
```
